---
title: "Graduate School Admissions Data Report"
author: "Erick Alvarez"
date: "December 2, 2024"
output: 
    html_document:
        theme: paper
        highlight: tango
        toc: true
        toc_float:
            collapsed: true
        number_sections: false
        code_download: true
        df_print: kable
        code_folding: show
        mode: selfcontained
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, warning = FALSE, message = FALSE)

```

```{r, echo = FALSE}
# load all necessary libraries within this code chunk 
library(tidyverse)
library(reactable)
library(GGally)
library(skimr)

# converts all numbers to standard notation 
options(scipen = 999)

```

# Data Set
The data set being examined consists of graduate school admissions. The CSV file being read in is titled "GradSchool_Admissions.csv" as found in the 'Data' folder in the BIOL 3100 GitHub repository. A portion of the data frame is seen below: 

```{r, echo = FALSE}
# read in the data 
grad_dat <- read.csv("../Data/GradSchool_Admissions.csv")

# show the first 6 rows of the data 
head(grad_dat)

```
The data consists of 4 columns:

1. **admit** 
    - Reflects the acceptance or rejection of a student's application 
      - 1 = Successful (Admitted)
      - 0 = Failure (Rejected)
      
2. **gre**
    - Reflects the GRE (Graduate Record Examination) score of the individual student 
    
3. **gpa** 
    - Reflects the GPA (Grade Point Average) out of 4.0 of the individual student 
    
4. **rank** 
    - Reflects the overall ranking of the student's undergraduate institution 
      - Rank of 1 is "Top Tier"
      - Lowest rank is 4 
      
Here is the data structure itself: 

```{r, echo = FALSE}
# show the structure of the data frame and its data types 
str(grad_dat)

```

# Exploratory Data Analysis 
I know I want to model admittance outcome predictions according to a student's institutional rank, GPA, and GRE score. However, I first need to determine which of the predictor variables has the strongest/weakest correlation to the response variable. 

The following sections consist of the exploratory data analyses pertaining to these correlations. 

## Distribution of Predictor Variables 
Firstly, I will be investigating the distribution of the predictor variables on their own so I can infer any hidden trends or patterns that may be important in modeling the admittance predictions. 

The following graph is a histogram of the GRE scores across the entire data set:

```{r, echo = FALSE}
grad_dat |>
  ggplot(aes(x = gre)) +
    geom_histogram(binwidth = 20, fill = "lightblue", color = "blue") +
    labs(title = "Histogram of GRE Scores", x = "GRE Score", y = "Number of Students") +
    theme_minimal()

```

The following table also contains the summary statistics of the same GRE distribution:

```{r, echo = FALSE}
# calculate summary statistics for GRE scores
gre_summary_stats <- grad_dat |>
  summarize(
    Mean = mean(gre, na.rm = TRUE),
    Median = median(gre, na.rm = TRUE),
    SD = sd(gre, na.rm = TRUE),
    Min = min(gre, na.rm = TRUE),
    Max = max(gre, na.rm = TRUE),
    Range = max(gre, na.rm = TRUE) - min(gre, na.rm = TRUE),
    Q1 = quantile(gre, 0.25, na.rm = TRUE),
    Q3 = quantile(gre, 0.75, na.rm = TRUE),
    IQR = IQR(gre, na.rm = TRUE)
  )

# create a reactable table for the summary statistics
reactable(gre_summary_stats, 
          columns = list(
            Mean = colDef(name = "Mean"),
            Median = colDef(name = "Median"),
            SD = colDef(name = "Standard Deviation"),
            Min = colDef(name = "Min"),
            Max = colDef(name = "Max"),
            Range = colDef(name = "Range"),
            Q1 = colDef(name = "25th Percentile (Q1)"),
            Q3 = colDef(name = "75th Percentile (Q3)"),
            IQR = colDef(name = "Interquartile Range (IQR)")
          ),
          theme = reactableTheme(
            color = "#2a3d47", 
            borderColor = "#e0e0e0", 
            highlightColor = "#f7f7f7"
          )
)

```

Next is a density plot showing the distribution of the GPA values. I used a density plot since we are dealing with float values instead of whole integers:

```{r, echo = FALSE}
grad_dat |> 
  ggplot(aes(x = gpa)) +
    geom_density(fill = "pink", color = "red", alpha = 0.5) +
    labs(title = "Density Plot of GPA", x = "GPA", y = "Density") +
    theme_minimal()

```

The following table also contains the summary statistics of the same GPA distribution:

```{r, echo = FALSE}
# calculate summary statistics for GPA scores
gpa_summary_stats <- grad_dat |>
  summarize(
    Mean = mean(gpa, na.rm = TRUE),
    Median = median(gpa, na.rm = TRUE),
    SD = sd(gpa, na.rm = TRUE),
    Min = min(gpa, na.rm = TRUE),
    Max = max(gpa, na.rm = TRUE),
    Range = max(gpa, na.rm = TRUE) - min(gpa, na.rm = TRUE),
    Q1 = quantile(gpa, 0.25, na.rm = TRUE),
    Q3 = quantile(gpa, 0.75, na.rm = TRUE),
    IQR = IQR(gpa, na.rm = TRUE)
  )

# create a reactable table for the summary statistics
reactable(gpa_summary_stats, 
          columns = list(
            Mean = colDef(name = "Mean"),
            Median = colDef(name = "Median"),
            SD = colDef(name = "Standard Deviation"),
            Min = colDef(name = "Min"),
            Max = colDef(name = "Max"),
            Range = colDef(name = "Range"),
            Q1 = colDef(name = "25th Percentile (Q1)"),
            Q3 = colDef(name = "75th Percentile (Q3)"),
            IQR = colDef(name = "Interquartile Range (IQR)")
          ),
          theme = reactableTheme(
            color = "#2a3d47", 
            borderColor = "#e0e0e0", 
            highlightColor = "#f7f7f7"
          )
)

```

Lastly, here is the distribution of undergraduate institutional rankings: 

```{r, echo = FALSE}
grad_dat |>
  ggplot(aes(x = factor(rank))) +  # Use 'factor' to treat rank as a categorical variable
    geom_bar(fill = "darkgreen", color = "lightgreen") +
    labs(title = "Distribution of School Ranks", x = "Rank", y = "Number of Students") +
    theme_minimal() +
    annotate("text", x = 2.6, y = 60, label = "1 = Top-Tier\n4 = Low-Tier", color = "black", size = 5, hjust = 0)

```

I'm not including any summary statistics for the school rank as any important inference that could be made with categorical data can be easily seen through the graphed visual. 

